Proportional multi-state multiple-cohort life table model

### 2018-04-09 MSLT/MSLTdoc.Rmd
Spell check

Data source: (1) National census; (2) Worldwide population and mortality data:  http://www.mortality.org/ (mostly high income countries; and (3) Calculate from the Global Buden of Disease by the Institute of Health Metrics and Evaluation (GBD IHME) data (rates and numbers available from (http://ghdx.healthdata.org/gbd-results-tool).

@@

Note that we need data for population numbers and all cause mortality rates for: (1) PMSLT and (2) Dismod II collection (more in Dismod II section). Population data from the synthetic population is used for the PMSLT. For Dismod II, popuation and mortality data should be from the same source (GBD IHME)

@@

We start by cleaning the global environment (#1) to keep track of our works and ensure that the code is generating our outcomes. Then, we set up an option to avoid the use of scientific notation (#2) and lastly we load the functions (#3). The code chunks are shown in the rmarkdown ouput.

@@

Initial case studies for the ITHIMR are: London, Sao PAblo, Delhi, Accra, Los Angeles and Edinburgh. Here, we will start with **greater London** given the availability of dieases epidemiology data from the GBD IHME study. For the rest of the case study cities data is available at the country level, hence, a scaling method is needed.

@@

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**. The color coding from Data inputs sources to Inputs PMSLT link sources with inputs for the PMSLT. Solid arrows represent final inputs and dashed-arrows represent intermediate inputs that need further processing. Purple coding means a process and green coding represent change in mortality and disability prevalence rates to modify the life table parameters. Black color coding with white color font represent final model outcomes. For both, the life table and disease life table, two sets of each are simulated, one for the baseline and the other for the scenario. **Data inputs** are from four sources: Sythetic population, GBD data (from the Institute of Health Metrics and Evaluations (IHME)), ITHIM calculations and Injuries model. The synthetic population provides population numbers by age-groups and sex which are used in the life table (**PMSLT Mechanisms**) with each age cohort simulated until they die or reach 100 years of age. Population numbers are also used to derive road death rates and road injuries rates from numbers for the baseline and scenario generated by the Injuries model. The difference in road deaths rates modifies all cause-mortatliy in the scenario life table and the change in injuries rates modifies total YLDs rates (hence, healthy life years). Data from GBD IHME includes: total YLD rates per one, disease specific YLDs (numbers) and prevalence to derive disability weights for the included diseases, all cause mortatliy rates per one, population numbers and diseases parameters (incidence, mortality and prevalence). Dismod II is used to derive internally consitent estimates incidence and case fatality rates from GBD IHME data, which are inputs of the disease specific life tables. Case fatality derive from Dismod II is the same for the baseline and scenario (however, could be changed) and incidence is different. For the baseline life tables incidence is as derived from Dismod II and for the scanrio is modified by the PIF (1-PIF). A change in incidence modifies prevalence and mortality in later years. The difference between prevalence and mortality rates between the baseline and scanrio then feed into the scenario life table.*](structure\Figure3.png){width=50%,hight=30%,fig.pos="H"}

@@

![Global Burden of Disease data results tool.](data/Figure4.png){width=70%,hight=50%,fig.pos="H"}

@@

@@

(NEED TO PREPARE: LONDON, SAO PABLO, DELHI, ACCRA, LOS ANGELES, EDINBURGH)). DATA AVAILABILITY FROM ARE IN 5 YEARS UP TO 95+. gbd DATA IS NOT AVAILABLE AT THE CITY LEVEL, HOWEVER, SOME LEVEL OF DISSAGREAGATION IS AVAILABLE. ENGLAND: EAST MIDLANDS, EAST OF ENGLAND, GRETER LONDON, ETC. wE COULD USE GREATER LONDON. FOR EDINBURGH WE HAVE DATA FOR SCOTLAND.  FOR SAO PABLO, DATA IS FOR BRAZIL. MEXICO IS NOT LISTED, BUT GBD HAS DATA FOR DISTRITO FEDERAL. FOR DELHI THE DATA IS FOR INDIA. FOR ACCRA THE DATA IS FOR GHANA. FOR LOS ANGELES DATA IS FOR UNITED STATES, BUT WE CAN HAVE THIS FROM NEIL?

The code below is to load the data used will be updated to the study cases.

WRITE CODE HERE TO GET DATA.

@@

We define parameters for our model that are used multiple times. *p_age_cohort* defines the starting age for the cohorts. *p_sex* is for the cohort sex. *p_disease* specifies the modelled diseases. *p_intervention* is a hypothetical increase in MET minutes per week.

@@

The following code applies to the data used as an exmaple here. #6 transforms upper cases to lower cases in the sex variable. #7 and #8 is used to rename mortality_rate to mx (used in life table language). #9 is used to replace male and female in the exposure data to males and females to facilitate combination of data sets.

>>>>

Data source: (1) National census; (2) Worldwide population and mortality data:  http://www.mortality.org/ (mostly high income countries; and (3) Calculate from the Global Burden of Disease by the Institute of Health Metrics and Evaluation (GBD IHME) data (rates and numbers available from (http://ghdx.healthdata.org/gbd-results-tool).

@@

Note that we need data for population numbers and all cause mortality rates for: (1) PMSLT and (2) Dismod II collection (more in Dismod II section). Population data from the synthetic population is used for the PMSLT. For Dismod II, population and mortality data should be from the same source (GBD IHME)

@@

We start by cleaning the global environment (#1) to keep track of our works and ensure that the code is generating our outcomes. Then, we set up an option to avoid the use of scientific notation (#2) and lastly we load the functions (#3). The code chunks are shown in the rmarkdown output.

@@

Initial case studies for the ITHIMR are: London, Sao Pablo, Delhi, Accra, Los Angeles and Edinburgh. Here, we will start with **greater London** given the availability of diseases epidemiology data from the GBD IHME study. For the rest of the case study cities data is available at the country level, hence, a scaling method is needed.

@@

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**. The color coding from Data inputs sources to Inputs PMSLT link sources with inputs for the PMSLT. Solid arrows represent final inputs and dashed-arrows represent intermediate inputs that need further processing. Purple coding means a process and green coding represent change in mortality and disability prevalence rates to modify the life table parameters. Black color coding with white color font represent final model outcomes. For both, the life table and disease life table, two sets of each are simulated, one for the baseline and the other for the scenario. **Data inputs** are from four sources: Sythetic population, GBD data (from the Institute of Health Metrics and Evaluations (IHME)), ITHIM calculations and Injuries model. The synthetic population provides population numbers by age-groups and sex which are used in the life table (**PMSLT Mechanisms**) with each age cohort simulated until they die or reach 100 years of age. Population numbers are also used to derive road death rates and road injuries rates from numbers for the baseline and scenario generated by the Injuries model. The difference in road deaths rates modifies all cause-mortatliy in the scenario life table and the change in injuries rates modifies total YLDs rates (hence, healthy life years). Data from GBD IHME includes: total YLD rates per one, disease specific YLDs (numbers) and prevalence to derive disability weights for the included diseases, all cause mortatliy rates per one, population numbers and diseases parameters (incidence, mortality and prevalence). Dismod II is used to derive internally consitent estimates incidence and case fatality rates from GBD IHME data, which are inputs of the disease specific life tables. Case fatality derive from Dismod II is the same for the baseline and scenario (however, could be changed) and incidence is different. For the baseline life tables incidence is as derived from Dismod II and for the scanrio is modified by the PIF (1-PIF). A change in incidence modifies prevalence and mortality in later years. The difference between prevalence and mortality rates between the baseline and scanrio then feed into the scenario life table.*](structure\Figure3.png){width=50%,height=30%,fig.pos="H"}

@@

![Global Burden of Disease data results tool.](data/Figure4.png){width=70%,height=50%,fig.pos="H"}

@@

Once the selections described in Table 2 are made, the option *Download CVS** in the GHDx website is selected. A prompt comes up asking for an email address. The data is sent to the designated email address (within minutes) in ZIP format, unzip and use the code below to read the data. Here, we selected data for Greater London and England. The aim is to compare and derive scaling factors as for most cities the data is not available from the GBD and country level data may be used and scaled to the city level. Note that the data input requirement for the PMSLT, except population numbers, is in rates. Therefore the scaling is to better reflect the burden of an area, this is a different issue than working with numbers (e.g. total mortality numbers, total YLDs numbers) as in the ITHIM approach.

```{r, echo=TRUE}
options(scipen=999)
GBDdata <- read.csv("data/IHME-GBD_2016_DATA-8fca12f6-1.csv", stringsAsFactors = F)
```

@@

@@

We define parameters for our model that are used multiple times. *p_age_cohort* defines the starting age for the cohorts. *p_sex* is for the cohort sex. *p_disease* specifies the modeled diseases. *p_intervention* is a hypothetical increase in MET minutes per week.

@@

The following code applies to the data used as an example here. #6 transforms upper cases to lower cases in the sex variable. #7 and #8 is used to rename mortality_rate to mx (used in life table language). #9 is used to replace male and female in the exposure data to males and females to facilitate combination of data sets.
